The Weakly Nonlinear Magnetorotational Instability

### 2016-08-31 paper/widegap_mri_paper.tex
begin long run at Pm=1.6E-5

a Ginzburg-Landau equation (GLE). The GLE governs the weakly nonlinear amplitude behavior in a wide range of physical systems, including the narrow gap MRI (\citei{Umurhan:2007hs}), Rayleigh-B\'enard convection, and hydrodynamic Taylor Couette flow. We emphasize that this is a model equation, valid only near marginality (e.g. \citei{Cross:1993el}). The dynamics of the GLE are determined by its coefficients, which are in turn determined by the linear eigenfunctions and nonlinear vectors plotted in Figures \ref{fig:linear_eigenfunctions} and \ref{fig:N2_N3}. Equation \ref{eq:gle} contains three coefficients: $b$, which determines the linear growth rate of the system, $d$, a diffusion coefficient, and $c$, the coefficient of the nonlinear term. When all of the coefficients of Equation \ref{eq:gle} are real, this is known as the real GLE, although the amplitude $\alpha$ remains complex. The real GLE is subject to several well-studied instabilities, including the Ekhaus and Zig-Zag instabilities. When the coefficients are complex, we have the complex GLE, a source of even richer phase dynamics than its real counterpart (e.g. \citei{Aranson:2002}).

@@

The preservation of symmetries in the thin-gap geometry is worth a closer look, as its absence in the wide gap case is the source of many differences in the systems. \citei{Latter:2015} point out that in the ideal limit ($\nu, \eta \rightarrow 0$), the linearized system described by the lefthand side of Equations \ref{eq:Psi_perturbed} - \ref{eq:Bphi_perturbed} can be expressed as a Shr{\"o}dinger equation for the radial velocity. Similarly combining equations to obtain a single expression for $\Psi$, we find that the thin-gap limit linear ideal MRI can be expressed as

@@

Our work should be placed in the broader context of emergent pattern formation in physical systems. The real Ginzburg-Landau equation derived here governs the slow-parameter evolution of the standard MRI close to threshold. The GLE arises in a number of other physical systems, and in each case it is a consequence not of the particular physics at hand, but of the underlying symmetries in the problem. Here we make a brief comparison to two other systems that give rise to a GLE. The first and perhaps most famous is Rayleigh-B\'enard convection, in which a fluid between two plates is heated from below. If we take the plane of the fluid to be infinite in the horizontal plane, the system is initially translationally symmetric. %When the system eventually becomes unstable to convective heat transfer it undergoes a symmetry breaking, allowing the 
At the onset of convection the system undergoes a symmetry breaking, forming rolls, or convection cells, which break the horizontal translational invariance. Analogously, the standard MRI system considered here is initially vertically translationally symmetric, because we idealize the Taylor-Couette device as an infinitely long cylinder. The MRI breaks this symmetry, forming cells along the vertical length of the domain. Just as Rayleigh-B\'enard cells transport heat vertically, the MRI cells transport angular momentum horizontally. The symmetry breaking of each of these systems is described near onset by the real GLE.

A real GLE has also been found to describe the formation of zonal flows out of magnetized turbulence in a model system (). Zonal flows are axisymmetric structures, large-scale and long-lived, which form spontaneously out of turbulence. They have recently been observed in some numerical studies of the MRI, and have generated considerable interest for their possible role in planet formation in protoplanetary disks (). The present work is of course an idealized geometry, and we make no attempt to model a realistic protoplanetary disk environment. However, it is worth noting that the GLE we derive implies that axisymmetric, large-scale, long-lived structures are a generic feature of the MRI in the weakly nonlinear regime. This work provides a mathematical description of the MRI as a pattern-forming process, but much remains to be understood, particularly involving the relevance of these insights to realistic astrophysical disks. Our model is most directly relevant to Taylor-Couette flows, and we emphasize that laboratory MRI experiments stand poised to observe the MRI-driven pattern formation predicted here.

@@

@@

@@

\bibitem[Pessah(2010)]{Pessah:2010ic}
Pessah, M, 2010, ApJ, 716, 1012

>>>>

a Ginzburg-Landau equation (GLE). The GLE governs the weakly nonlinear amplitude behavior in a wide range of physical systems, including the narrow gap MRI (\citei{Umurhan:2007hs}), Rayleigh-B\'enard convection (\citei{Newell:1969wr}), and hydrodynamic Taylor Couette flow (e.g. \citei{Recktenwald:1993}). We emphasize that this is a model equation, valid only near marginality (e.g. \citei{Cross:1993el}). The dynamics of the GLE are determined by its coefficients, which are in turn determined by the linear eigenfunctions and nonlinear vectors plotted in Figures \ref{fig:linear_eigenfunctions} and \ref{fig:N2_N3}. Equation \ref{eq:gle} contains three coefficients: $b$, which determines the linear growth rate of the system, $d$, a diffusion coefficient, and $c$, the coefficient of the nonlinear term. When all of the coefficients of Equation \ref{eq:gle} are real, this is known as the real GLE, although the amplitude $\alpha$ is in general complex. The real GLE is subject to several well-studied instabilities, including the Ekhaus and Zig-Zag instabilities. When the coefficients are complex, we have the complex GLE, a source of even richer phase dynamics than its real counterpart (see \citei{Aranson:2002} for a thorough review).

@@

The preservation of symmetries in the thin-gap geometry is worth a closer look, as its absence in the wide gap case is the source of many differences in the systems. \citei{Latter:2015} point out that in the ideal limit ($\nu, \eta \rightarrow 0$), the linearized system described by the lefthand side of Equations \ref{eq:Psi_perturbed} - \ref{eq:Bphi_perturbed} can be expressed as a Schr{\"o}dinger equation for the radial velocity. Similarly combining equations to obtain a single expression for $\Psi$, we find that the thin-gap limit linear ideal MRI can be expressed as

@@

Our work should be placed in the broader context of emergent pattern formation in physical systems. The real Ginzburg-Landau equation derived here governs the slow-parameter evolution of the standard MRI close to threshold. The GLE arises in a number of other physical systems, and in each case it is a consequence not of the particular physics at hand, but of the underlying symmetries in the problem. Here we make a brief comparison to two other systems that give rise to a GLE. The first and perhaps most famous is Rayleigh-B\'enard convection, in which a fluid between two plates is heated from below (\citei{Newell:1969wr}). If we take the plane of the fluid to be infinite in the horizontal plane, the system is initially translationally symmetric. %When the system eventually becomes unstable to convective heat transfer it undergoes a symmetry breaking, allowing the 
At the onset of convection the system undergoes a symmetry breaking, forming rolls, or convection cells, which break the horizontal translational invariance. Analogously, the standard MRI system considered here is initially vertically translationally symmetric, because we idealize the Taylor-Couette device as an infinitely long cylinder. The MRI breaks this symmetry, forming cells along the vertical length of the domain. Just as Rayleigh-B\'enard cells transport heat vertically, the MRI cells transport angular momentum horizontally. The symmetry breaking of each of these systems is described near onset by the real GLE.

A real GLE has also been found to describe the formation of zonal flows out of magnetized turbulence in a model system (\citei{Parker:2013hy}). Zonal flows are axisymmetric structures, large-scale and long-lived, which form spontaneously out of turbulence. They have recently been observed in some numerical studies of the MRI, and have generated considerable interest for their possible role in planet formation in protoplanetary disks (\citei{Johansen:2009uj}, \citei{Kunz:2013}). The present work is of course an idealized geometry, and we make no attempt to model a realistic protoplanetary disk environment. However, it is worth noting that the GLE we derive implies that axisymmetric, large-scale, long-lived structures are a generic feature of the MRI in the weakly nonlinear regime. This work provides a mathematical description of the MRI as a pattern-forming process, but much remains to be understood, particularly involving the relevance of these insights to realistic astrophysical disks. Our model is most directly relevant to Taylor-Couette flows, and we emphasize that laboratory MRI experiments stand poised to observe the MRI-driven pattern formation predicted here.

\section{Acknowledgements}
NASA, NSF. The authors thank Mordecai Mac Low, Jeremy Goodman, etc for useful discussion.

@@

\bibitem[{Johansen et al.(2009)}]{Johansen:2009uj}
Johansen, A., Youdin, A., Klahr, H., 2009, ApJ 697, 1269

@@

\bibitem[{Kunz \& Lesur(2013)}]{Kunz:2013}
Kunz, M.W. \& Lesur, G., 2013, MNRAS 434, 2295

@@

\bibitem[{Newell \& Whitehead(1969)}]{Newell:1969wr}
Newell, A.C. \& Whitehead, J.A., 1969, JFM 38, 279

\bibitem[{Parker \& Krommes(2013)}]{Parker:2013hy}
Parker, J.B. \& Krommes, J.A., 2013, Phys of Plasmas 20, 100703

\bibitem[{Pessah(2010)}]{Pessah:2010ic}
Pessah, M, 2010, ApJ, 716, 1012

\bibitem[{Recktenwald et al.(1993)}]{Recktenwald:1993}
Recktenwald, A., L\"ucke, M., M\"uller, H.W., 1993, Phys Rev E 48, 4444
